Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-16 manuscript/manuscript.tex
Remove references to "coarse graining"

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
This suggests a binned SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.

We can also compute the pointwise mutual information in this coarse-grained distribution as
\begin{align}
    \hilopmi(d; i_c) = \log \frac
                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
\end{align}
One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.
Also, it is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.

@@

On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, $D = -0.42$) also generates positive $\fpmi$.
The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
This justifies coarse-graining the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.

Accordingly, we computed the coarse-grained $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.

@@

\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our coarse-grained $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

Our results above show that $\fpmi$ and its coarse-grained analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

>>>>

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
This suggests a binned SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
It is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.

We can also compute the pointwise mutual information in this binned distribution as
\begin{align}
    \hilopmi(d; i_c) = \log \frac
                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
\end{align}
One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.

@@

On the other hand, a beta coalescent that generates similar distortions in the SFS ($\alpha=1.45$, $D = -0.42$) also generates positive $\fpmi$.
The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
This justifies binning the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.

Accordingly, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.

@@

\Fig{fig:wpmi_recombination} shows the weighted $\fpmi$ for three genetic distances in a constant-$N$ Kingman coalescent, a Kingman coalescent with exponential growth, and a beta coalescent.
As in the non-recombining case, the constant-$N$ and exponential-growth Kingman models have similar $\fpmi$.
In both models, $\fpmi > 0$ for $i\approx j$ for $r\E{T_2} \sim 1$.
This is presumably because trees at nearby sites contain clades with similar numbers of leaves.
These positive correlations do not extend to $i \gg j$, which is the signal of multiple mergers in our binned $\hilopmi$.
On the other hand, the positive $\fpmi$ in the beta coalescent persists for $r\E{T_2} > 1$.
Thus, the signal of multiple mergers in $\fpmi$ is robust to recombination.

@@

Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.
